---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# spacelaser

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Cloud-optimized partial reading of GEDI and ICESat-2 HDF5 data from R.
Only the bytes needed for the requested spatial subset are fetched over
HTTP, avoiding multi-gigabyte downloads.

Supports GEDI L1B, L2A, L2B, L4A, L4C and ICESat-2 ATL03, ATL06, ATL07,
ATL08, ATL10, ATL13, ATL24.

## Installation

Requires a [Rust toolchain](https://www.rust-lang.org/tools/install)
(cargo + rustc).

``` r
# install.packages("pak")
pak::pak("belian-earth/spacelaser")
```

## Authentication

NASA Earthdata credentials are resolved from `EARTHDATA_USERNAME` /
`EARTHDATA_PASSWORD` or a `~/.netrc` entry for `urs.earthdata.nasa.gov`.
Register at <https://urs.earthdata.nasa.gov/> if you don't have an
account. The simplest setup:

```{r auth, eval = FALSE}
earthdatalogin::edl_netrc()
```

## Example

```{r example}
library(spacelaser)

bbox <- sl_bbox(-124.04, 41.39, -124.01, 41.42)

# GEDI
granules <- sl_search(
  bbox, 
  product = "L2A", 
  date_start = "2022-01-01", 
  date_end = "2023-01-01")
gedi2a <- sl_read(granules)

gedi2a

# # ICESat-2 uses the same two verbs
# granules <- sl_search(bbox, product = "ATL08")
# icesat <- sl_read(granules)
```

Columns available for a product:

```{r cols, eval = TRUE}
sl_columns("L2A")
```


