---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# spacelaser

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![extendr](https://img.shields.io/badge/extendr-*-276DC2)](https://extendr.github.io/extendr/extendr_api/)
[![License:Apache](https://img.shields.io/github/license/belian-earth/a5R)](https://www.apache.org/licenses/LICENSE-2.0)
[![Codecov test coverage](https://codecov.io/gh/belian-earth/spacelaser/graph/badge.svg)](https://app.codecov.io/gh/belian-earth/spacelaser)
<!-- badges: end -->

Cloud-optimized partial reading of GEDI and ICESat-2 HDF5 data from R.
Only the bytes needed for the requested spatial subset are fetched over
HTTP, avoiding multi-gigabyte downloads.

## Supported products

### GEDI

| Product | Description                                   |
|:--------|:----------------------------------------------|
| L1B     | Geolocated received and transmitted waveforms |
| L2A     | Elevation and canopy relative height metrics  |
| L2B     | Canopy cover fraction and vertical profile    |
| L4A     | Footprint-level aboveground biomass density   |
| L4C     | Waveform structural complexity index          |

### ICESat-2

| Product | Description                          |
|:--------|:-------------------------------------|
| ATL03   | Geolocated photon heights            |
| ATL06   | Land ice surface elevation           |
| ATL07   | Sea ice surface elevation            |
| ATL08   | Land and vegetation height segments  |
| ATL10   | Sea ice freeboard                    |
| ATL13   | Inland water surface heights         |
| ATL24   | Near-shore bathymetric photons       |

## Installation

Requires a [Rust toolchain](https://www.rust-lang.org/tools/install)
(cargo + rustc).

``` r
# install.packages("pak")
pak::pak("belian-earth/spacelaser")
```

## Authentication

All reads go through NASA Earthdata, which requires a free account.
Register at <https://urs.earthdata.nasa.gov/>.

Credentials can be supplied in any of the following ways:

- **Environment variables** — set `EARTHDATA_USERNAME` and
  `EARTHDATA_PASSWORD`. Convenient for CI and shell sessions.
- **A `.netrc` file** — add an entry for `urs.earthdata.nasa.gov` to
  `~/.netrc` (or `_netrc` on Windows). spacelaser will read it directly.
- [**`earthdatalogin`**](https://boettiger-lab.github.io/earthdatalogin/) — the simplest option if you don't already have a
  netrc set up:

```{r auth, eval = FALSE}
# install.packages("earthdatalogin")
earthdatalogin::edl_netrc()
```

This writes a netrc for you and is interoperable with other R
Earthdata tools.

## Example

```{r example, eval=FALSE}
library(spacelaser)

bbox <- sl_bbox(-124.04, 41.39, -124.01, 41.42)

# GEDI
granules <- sl_search(
  bbox, 
  product = "L2A", 
  date_start = "2022-01-01", 
  date_end = "2023-01-01")
gedi2a <- sl_read(granules)

gedi2a

# # ICESat-2 uses the same two verbs
# granules <- sl_search(bbox, product = "ATL08")
# icesat <- sl_read(granules)
```

Columns available for a product:

```{r cols, eval = FALSE}
sl_columns("L2A")
```


