---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# spacelaser

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![extendr](https://img.shields.io/badge/extendr-*-276DC2)](https://extendr.github.io/extendr/extendr_api/)
[![License:Apache](https://img.shields.io/github/license/belian-earth/a5R)](https://www.apache.org/licenses/LICENSE-2.0)
[![Codecov test coverage](https://codecov.io/gh/belian-earth/spacelaser/graph/badge.svg)](https://app.codecov.io/gh/belian-earth/spacelaser)
[![R-CMD-check](https://github.com/belian-earth/spacelaser/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/belian-earth/spacelaser/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Cloud-optimized partial reading of GEDI and ICESat-2 HDF5 data from R.
Only the bytes needed for the requested spatial subset are fetched over
HTTP, avoiding multi-gigabyte downloads.

## Supported products

### GEDI

| Product | Description                                   |
|:--------|:----------------------------------------------|
| L1B     | Geolocated received and transmitted waveforms |
| L2A     | Elevation and canopy relative height metrics  |
| L2B     | Canopy cover fraction and vertical profile    |
| L4A     | Footprint-level aboveground biomass density   |
| L4C     | Waveform structural complexity index          |

### ICESat-2

| Product | Description                          |
|:--------|:-------------------------------------|
| ATL03   | Geolocated photon heights            |
| ATL06   | Land ice surface elevation           |
| ATL07   | Sea ice surface elevation            |
| ATL08   | Land and vegetation height segments  |
| ATL10   | Sea ice freeboard                    |
| ATL13   | Inland water surface heights         |
| ATL24   | Near-shore bathymetric photons       |

## Why spacelaser

The standard R workflow for GEDI / ICESat-2 data is to download whole
HDF5 granules, then filter locally. For a typical spatial subset query
that wastes minutes and gigabytes — the file you're filtering is
usually 100-10,000× larger than the answer you actually want.

Spacelaser issues HTTP Range requests against the remote files and
returns just the rows that fall inside your bounding box, with no
local caching needed.

Cold-cache benchmark, GEDI L2A, 0.03° × 0.03° bbox over Gabon,
2 years, 11 granules, ecologist-realistic column set:

| | spacelaser | curl + hdf5r (status quo) |
|---|---|---|
| Wall time | **70 s** | 1,568 s |
| Bytes downloaded | (small) | 27 GiB |
| Disk used | 0 | 27 GiB |
| Output | 1,246 shots × 112 cols | 1,246 shots × 112 cols (bit-perfect equivalent) |

That's a **22.5×** speedup, robustly **15-22×** across runs depending
on NASA network conditions. Full methodology, equivalence-check
machinery, reproduction instructions, and all archived runs in
[`benchmarks/`](benchmarks/).

## Installation

Requires a [Rust toolchain](https://www.rust-lang.org/tools/install)
(cargo + rustc).

``` r
# install.packages("pak")
pak::pak("belian-earth/spacelaser")
```

## Authentication

All reads go through NASA Earthdata, which requires a free account.
Register at <https://urs.earthdata.nasa.gov/>.

Credentials can be supplied in any of the following ways:

- **Environment variables** — set `EARTHDATA_USERNAME` and
  `EARTHDATA_PASSWORD`. Convenient for CI and shell sessions.
- **A `.netrc` file** — add an entry for `urs.earthdata.nasa.gov` to
  `~/.netrc` (or `_netrc` on Windows). spacelaser will read it directly.
- [**`earthdatalogin`**](https://boettiger-lab.github.io/earthdatalogin/) — the simplest option if you don't already have a
  netrc set up:

```{r auth, eval = FALSE}
# install.packages("earthdatalogin")
earthdatalogin::edl_netrc()
```

This writes a netrc for you and is interoperable with other R
Earthdata tools.

## Example

```{r example, eval=FALSE}
library(spacelaser)

bbox <- sl_bbox(-124.04, 41.39, -124.01, 41.42)

# GEDI
granules <- sl_search(
  bbox, 
  product = "L2A", 
  date_start = "2022-01-01", 
  date_end = "2023-01-01")
gedi2a <- sl_read(granules)

gedi2a

# # ICESat-2 uses the same two verbs
# granules <- sl_search(bbox, product = "ATL08")
# icesat <- sl_read(granules)
```

Columns available for a product:

```{r cols, eval = FALSE}
sl_columns("L2A")
```


