---
title: "GEDI L1B waveforms: from raw returns to height above ground"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GEDI L1B waveforms: from raw returns to height above ground}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Live chunks hit NASA CMR + DAAC. They run only when EARTHDATA_USERNAME
# is set — CRAN builds and credential-less contributors see the rendered
# narrative without triggering network or auth.
eval_live <- nzchar(Sys.getenv("EARTHDATA_USERNAME"))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup}
library(spacelaser)
suppressPackageStartupMessages(library(dplyr))
library(ggplot2)
```

## What's a GEDI L1B waveform, and why bother?

GEDI L2A gives you summary elevation metrics per shot — ground elevation,
relative height quantiles (RH0 through RH100), canopy energy totals. That's
what most users want most of the time. L1B is the step before: one full
*waveform* per shot, capturing the lidar return amplitude as a function
of elevation. The L1B waveform has already been smoothed with a Gaussian
of width 2 ns to remove ADC sampling artefacts (Hofton & Blair, 2019,
Section 3.2.3); it is not the raw telemetered signal (that lives in L1A,
which is not publicly distributed). Everything in L2A is derived from
L1B by peak detection, additional smoothing, and mode fitting.

Working directly with L1B is worth it when you want to:

- See canopy structure, not just metrics derived from it
- Detect features L2A's algorithms didn't capture (understory layers,
  near-ground returns in steep terrain, water-surface returns)
- Build your own metrics or validate L2A's ground detection against the
  L1B signal

This vignette walks from L1B waveforms to height-above-detected-ground
plots by joining L1B with L2A on `shot_number` and using L2A's
`elev_lowestmode` as the per-shot zero reference.

## Setup

We'll work over a small bbox on the Pacific Northwest coast — coastal
redwoods, 50–100 m tall — where the difference between "height above
ellipsoid", "height above SRTM", and "height above detected ground" is
dramatic enough to be visually obvious.

```{r bbox-dates}
bbox <- sl_bbox(-124.04, 41.39, -124.01, 41.42)

dates <- list(start = "2020-06-01", end = "2020-09-01")
```

## Search and read both products

Search is cheap, so we issue it twice — once per product — and `sl_read`
over the shared bbox. Restrict to a single granule per product to keep the
vignette snappy; remove the `[1L, ]` for the full pipeline.

```{r search-read, eval = eval_live}
g_l1b <- sl_search(bbox, product = "L1B",
                   date_start = dates$start, date_end = dates$end)
g_l2a <- sl_search(bbox, product = "L2A",
                   date_start = dates$start, date_end = dates$end)

l1b <- sl_read(g_l1b[1L, ], bbox = bbox)
l2a <- sl_read(g_l2a[1L, ], bbox = bbox)
```

`l1b` carries the waveform (`rxwaveform`, already Gaussian-smoothed at
the L1B processing stage) and its structural metadata (`elevation_bin0`,
`elevation_lastbin`, `rx_sample_count`), plus noise statistics, DEM
references, and per-shot metadata. `l2a` carries the
processed ground elevation (`elev_lowestmode`), quality flags
(`quality_flag`, `sensitivity`, `degrade_flag`), and the relative-height
quantiles (`rh0` through `rh100`).

## Join L1B with L2A ground elevations

`shot_number` is the shared key. Two granules of the same orbit over the
same bbox cover the same shots, so a plain inner join gives you waveforms
and their detected ground elevations side by side.

```{r join, eval = eval_live}
l1b_joined <- l1b |>
  inner_join(
    l2a |> select(shot_number, elev_lowestmode, quality_flag, sensitivity,
                  rh25, rh50, rh75, rh95, rh100),
    by = "shot_number"
  ) |>
  filter(
    quality_flag == 1L,  # L2A quality gate
    sensitivity > 0.9,   # only well-penetrated shots
    solar_elevation < 0  # night shots for cleaner returns
  )

nrow(l1b_joined)
```

Quality filtering matters: L1B will happily hand you waveforms from shots
L2A flagged as low-sensitivity (e.g. dense cloud cover, high solar
contamination). Keep only the shots where L2A trusts its own ground
detection.

## Extract waveforms, then re-reference to ground

`sl_extract_waveforms` knows how to subtract DEM references, but it doesn't
know about L2A ground. Extract in ellipsoidal space (the default) and subtract
`elev_lowestmode` per shot post-hoc.

```{r extract, eval = eval_live}
wf <- sl_extract_waveforms(
  l1b_joined,
  height_ref = "ellipsoid", 
  normalise_amplitude = "snr"   # noise-normalised so beams/channels compare
)

# per-sample ground reference, looked up by shot
ground_by_sample <- l1b_joined$elev_lowestmode[
  match(wf$shot_number, l1b_joined$shot_number)
]

wf$height_above_ground <- wf$elevation - ground_by_sample
```

`height_above_ground = 0` is now the detected ground return; positive is
canopy, negative is noise samples below ground where the RX window
continues.

We used `normalise_amplitude = "snr"` above, which subtracts the per-shot
noise mean and divides by the per-shot noise standard deviation. This puts
every shot on a common scale regardless of which digitiser channel captured
it (GEDI's channels have different baseline pedestals; without
normalisation, waveforms from different beams cluster at different
baselines). The alternative `"noise"` subtracts the mean but skips the
division, leaving amplitude in baseline-subtracted ADC counts; `"raw"` (the default)
returns the waveform as-is. SNR units are convenient for overlaying multiple shots because the noise
floor sits at zero with unit variance across all shots, regardless of beam
or channel. A value of 4 would mean "4 standard deviations above noise",
which provides an intuitive scale for distinguishing signal from background.

## Anatomy of a single waveform

Before stacking hundreds of shots into a composite, it's worth unpacking
what a single waveform actually contains. Pick a shot with a tall canopy
(`rh95` above 50 m) and plot its amplitude vs height, annotated with the
L2A-derived structural metrics.

```{r single-shot, eval = eval_live, fig.width = 5, fig.height = 9}
tall_shot <- l1b_joined |>
  arrange(desc(rh95)) |>
  slice(5L) |>  # 5th tallest rh95 (avoids possible outliers at the very top)
  pull(shot_number)

wf_one  <- wf |> filter(shot_number == tall_shot)
meta_one <- l1b_joined |> filter(shot_number == tall_shot)

ggplot(wf_one, aes(amplitude, height_above_ground)) +
  geom_path(colour = "#1d743eff", linewidth = 0.5) +
  geom_hline(yintercept = 0,             colour = "#d48282",
             linewidth = 0.4) +
  geom_hline(yintercept = meta_one$rh50, colour = "#8782d4",
             linewidth = 0.3, linetype = "dashed") +
  geom_hline(yintercept = meta_one$rh95, colour = "#8782d4",
             linewidth = 0.3, linetype = "dashed") +
  annotate("text", x = Inf, y = 0,
           label = "ground", hjust = 1.1, vjust = -0.4,
           colour = "#d48282", size = 3) +
  annotate("text", x = Inf, y = meta_one$rh50,
           label = "rh50", hjust = 1.1, vjust = -0.4,
           colour = "#8782d4", size = 3) +
  annotate("text", x = Inf, y = meta_one$rh95,
           label = "rh95 (canopy top)", hjust = 1.1, vjust = -0.4,
           colour = "#8782d4", size = 3) +
  coord_cartesian(ylim = c(-10, 120)) +
  labs(
    x = "Amplitude (noise σ units)",
    y = "Height above detected ground (m)",
    title = sprintf("Single GEDI L1B waveform — rh95 = %.0f m",
                    meta_one$rh95)
  ) +
  theme_minimal()
```

Things to notice:

- The **strong peak at 0 m** is where L2A detected ground — that's why
  `height_above_ground = 0` lines up with it. Before ground-referencing
  this peak sat at some ellipsoidal elevation; we shifted it to zero
  here.
- The **broader peak around `rh95`** is approximately the canopy top. `rh95` is
  defined so 95 % of the waveform energy (cumulative from ground up)
  has been returned by that height — you can see the waveform fall off
  sharply just above it.
- `rh50` marks the median of the return energy, usually mid-canopy.
  Between `rh50` and `rh95` is where most of the foliage signal lives.
- Anything **above `rh95`** is either sparse emergent canopy, atmospheric
  return, or noise — typically very low amplitude.

## Comparing two shots: short vs tall

The same machinery over a different shot looks very different.
Pick one shot with low `rh95` and compare it to the tall one above. 

```{r two-shots, eval = eval_live}
# tall_shot is already defined from the anatomy section above
short_shot <- l1b_joined |>
  filter(rh95 < 15) |> slice(1L) |> pull(shot_number)

wf_pair <- wf |>
  filter(shot_number %in% c(short_shot, tall_shot)) |>
  left_join(
    l1b_joined |> select(shot_number, rh50, rh95),
    by = "shot_number"
  ) |>
  mutate(
    label = sprintf("shot %s\nrh50 = %.0f m, rh95 = %.0f m",
                    shot_number, rh50, rh95)
  )

ggplot(wf_pair, aes(amplitude, height_above_ground)) +
  geom_path(colour = "#1d743eff", linewidth = 0.5) +
  geom_hline(yintercept = 0, colour = "#d48282", linewidth = 0.3) +
  facet_wrap(~ label) +
  coord_cartesian(ylim = c(-10, 120)) +
  labs(
    x = "Amplitude (noise σ units)",
    y = "Height above detected ground (m)"
  ) +
  theme_minimal()
```

The short shot (open ground or low vegetation) is dominated by a single
ground peak with almost no energy above a 10 metres. The tall shot
has clear canopy structure: a ground return at zero plus a broad
canopy-top return at 50-90 m. 

## Composite view across the bbox

Now that we know what one waveform looks like, let's stack them all. We
overlay the amplitude quantiles at each height so the composite reads as
a structural profile rather than spaghetti.

```{r composite, eval = eval_live, fig.width = 5, fig.height = 9}
wf_quantiles <- wf |>
  filter(height_above_ground >= 0, height_above_ground <= 200) |>
  mutate(h_bin = round(height_above_ground)) |>
  group_by(h_bin) |>
  summarise(
    q10 = quantile(amplitude, 0.10, na.rm = TRUE),
    q50 = median(amplitude,         na.rm = TRUE),
    q90 = quantile(amplitude, 0.90, na.rm = TRUE),
    .groups = "drop"
  )

ggplot() +
  geom_ribbon(
    data = wf_quantiles,
    aes(xmin = q10, xmax = q90, y = h_bin),
    fill = "#24964fff", alpha = 0.1, colour= "#888888ff", linewidth = 0.4
  ) +
  geom_path(
    data = wf,
    aes(amplitude, height_above_ground, group = shot_number), colour = "#24964fff", linewidth = 0.1, alpha=0.6
  ) +
  
  geom_path(
    data = wf_quantiles,
    aes(q50, h_bin),
    colour = "#175e32ff", linewidth = 0.8
  ) +
  geom_hline(yintercept = 0, colour = "#d48282", linewidth = 0.8) +
  coord_cartesian(ylim = c(0, 120)) +
  scale_x_continuous(trans = "pseudo_log") +
  labs(
    x = "Amplitude (noise σ units)",
    y = "Height above detected ground (m)",
    title    = "L1B waveforms over PNW coastal redwood",
    subtitle = "Individual shots (green paths), 10-90% amplitude range (ribbon), median (line)"
  ) +
  theme_minimal()
```

The **median line** is the "typical" waveform for this bbox: a sharp
peak at ground, a secondary peak near median canopy height, decaying
returns above it. The **10-90 % ribbon** shows shot-to-shot variation
— narrow bands mean consistent structure across the bbox, wide bands
mean heterogeneous canopy (edge shots, gaps, short vs tall mixed).

## Cross-check: compute your own RH metrics

RH*N* is the elevation at which *N*% of cumulative waveform energy (from
ground up) has been returned. With the waveforms zeroed on ground, we can
compute it in-place and compare against L2A.

```{r rh-custom, eval = eval_live}
# cumulative energy from ground up, per shot
rh_custom <- wf |>
  filter(height_above_ground >= 0) |>       # ignore below-ground noise
  group_by(shot_number) |>
  arrange(height_above_ground, .by_group = TRUE) |>
  mutate(cum_energy = cumsum(amplitude) / sum(amplitude)) |>
  summarise(
    rh50_ours = approx(cum_energy, height_above_ground, xout = 0.50)$y,
    rh95_ours = approx(cum_energy, height_above_ground, xout = 0.95)$y,
    .groups = "drop"
  )

comparison <- l1b_joined |>
  select(shot_number, rh50, rh95) |>
  inner_join(rh_custom, by = "shot_number")

ggplot(comparison, aes(rh95, rh95_ours)) +
  geom_abline(colour = "#d48282") +
  geom_point(alpha = 0.4) +
  coord_equal() +
  labs(x = "L2A rh95 (m)", y = "Custom rh95 (m)")
```

If the points hug the 1:1 line, our simple-minded cumulative-energy RH95
agrees with L2A's more principled version. Disagreements identify shots
where L2A's noise thresholding, smoothing, or algorithm selection
(`selected_algorithm`) produces a meaningfully different answer than a
naive cumsum. Those disagreements are *L2A doing the right thing*,
not the other way around ;).

## What we skipped (and why)

- **Algorithm selection.** L2A ships six algorithm variants per shot and
  picks one via `selected_algorithm`. Our cumulative-energy RH uses none of
  that machinery.
- **Ground detection from scratch.** We used L2A's `elev_lowestmode` rather
  than detecting ground ourselves. naive ground detection in dense
  canopy or steep terrain is tricky and likely to be worse than L2A's algorithm; when you need the ground peak, read L2A.
- **Noise mean is not independently corrected.** Per the ATBD (Section
  3.2.1), "no corrections are currently applied to the mean noise
  level; the value from the telemetered datapacket is recorded."
  The column name `noise_mean_corrected` refers to the noise mean
  *of the corrected (smoothed) waveform*, not to an adjusted mean.
  Only `noise_stddev_corrected` receives an active correction (for
  odd-even sampler offsets, Section 3.2.2). This distinction is
  immaterial for the subtraction and SNR normalisation we perform
  here: the mean and standard deviation are measured on the same
  smoothed waveform we analyse, so the units match.

## Further reading

- Hofton, M. A. & Blair, J. B. (2019). *Algorithm Theoretical Basis
  Document for GEDI Transmit and Receive Waveform Processing for L1
  and L2 Products*, v1.0. NASA Goddard Space Flight Center. Covers
  L1B waveform smoothing, noise correction, and the L2A/L2B
  algorithm chain referenced throughout this vignette.
