---
title: "GEDI L1B waveforms: from raw returns to height above ground"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GEDI L1B waveforms: from raw returns to height above ground}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Live chunks hit NASA CMR + DAAC. They run only when EARTHDATA_USERNAME
# is set — CRAN builds and credential-less contributors see the rendered
# narrative without triggering network or auth.
eval_live <- nzchar(Sys.getenv("EARTHDATA_USERNAME"))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup}
library(spacelaser)
library(dplyr)
library(ggplot2)
```

## What's a GEDI L1B waveform, and why bother?

GEDI L2A gives you summary elevation metrics per shot — ground elevation,
relative height quantiles (RH0 through RH100), canopy energy totals. That's
what most users want most of the time. L1B is the step before: one full
*waveform* per shot, capturing the raw lidar return amplitude as a function
of elevation. Everything in L2A is derived from L1B by peak detection, noise
filtering, and mode fitting.

Working directly with L1B is worth it when you want to:

- See canopy structure, not just metrics derived from it
- Detect features L2A's algorithms didn't capture (understory layers,
  near-ground returns in steep terrain, water-surface returns)
- Build your own metrics or validate L2A's ground detection against the
  raw signal

This vignette walks from raw L1B waveforms to height-above-detected-ground
plots by joining L1B with L2A on `shot_number` and using L2A's
`elev_lowestmode` as the per-shot zero reference.

## Setup

We'll work over a small bbox on the Pacific Northwest coast — coastal
redwoods, 50–100 m tall — where the difference between "height above
ellipsoid", "height above SRTM", and "height above detected ground" is
dramatic enough to be visually obvious.

```{r bbox-dates}
bbox <- sl_bbox(-124.04, 41.39, -124.01, 41.42)

dates <- list(start = "2020-06-01", end = "2020-09-01")
```

## Search and read both products

Search is cheap, so we issue it twice — once per product — and `sl_read`
over the shared bbox. Restrict to a single granule per product to keep the
vignette snappy; remove the `[1L, ]` for the full pipeline.

```{r search-read, eval = eval_live}
g_l1b <- sl_search(bbox, product = "L1B",
                   date_start = dates$start, date_end = dates$end)
g_l2a <- sl_search(bbox, product = "L2A",
                   date_start = dates$start, date_end = dates$end)

l1b <- sl_read(g_l1b[1L, ], bbox = bbox)
l2a <- sl_read(g_l2a[1L, ], bbox = bbox)
```

`l1b` carries the raw waveform (`rxwaveform`) and its structural metadata
(`elevation_bin0`, `elevation_lastbin`, `rx_sample_count`), plus noise
statistics, DEM references, and per-shot metadata. `l2a` carries the
processed ground elevation (`elev_lowestmode`), quality flags
(`quality_flag`, `sensitivity`, `degrade_flag`), and the relative-height
quantiles (`rh0` through `rh100`).

## Join L1B with L2A ground elevations

`shot_number` is the shared key. Two granules of the same orbit over the
same bbox cover the same shots, so a plain inner join gives you waveforms
and their detected ground elevations side by side.

```{r join, eval = eval_live}
l1b_joined <- l1b |>
  inner_join(
    l2a |> select(shot_number, elev_lowestmode, quality_flag, sensitivity,
                  rh25, rh50, rh75, rh95, rh100),
    by = "shot_number"
  ) |>
  filter(
    quality_flag == 1L,  # L2A quality gate
    sensitivity > 0.9,   # only well-penetrated shots
    solar_elevation < 0  # night shots for cleaner returns
  )

nrow(l1b_joined)
```

Quality filtering matters: L1B will happily hand you waveforms from shots
L2A flagged as low-sensitivity (e.g. dense cloud cover, high solar
contamination). Keep only the shots where L2A trusts its own ground
detection.

## Extract waveforms, then re-reference to ground

`sl_extract_waveforms` knows how to subtract DEM references, but it doesn't
know about L2A ground. Extract in ellipsoidal space and subtract
`elev_lowestmode` per shot post-hoc.

```{r extract, eval = eval_live}
wf <- sl_extract_waveforms(
  l1b_joined,
  normalise_amplitude = "snr"   # baseline-aligned so beams/channels compare
)

# per-sample ground reference, looked up by shot
ground_by_sample <- l1b_joined$elev_lowestmode[
  match(wf$shot_number, l1b_joined$shot_number)
]

wf$height_above_ground <- wf$elevation - ground_by_sample
```

`height_above_ground = 0` is now the detected ground return; positive is
canopy, negative is noise samples below ground where the RX window
continues.

## Anatomy of a single waveform

Before stacking hundreds of shots into a composite, it's worth unpacking
what a single waveform actually contains. Pick a shot with a tall canopy
(`rh95` above 50 m) and plot its amplitude vs height, annotated with the
L2A-derived structural metrics.

```{r single-shot, eval = eval_live}
tall_shot <- l1b_joined |>
  arrange(desc(rh95)) |>
  slice(5L) |>  # top-5 avoids any outlier; adjust to taste
  pull(shot_number)

wf_one  <- wf |> filter(shot_number == tall_shot)
meta_one <- l1b_joined |> filter(shot_number == tall_shot)

ggplot(wf_one, aes(amplitude, height_above_ground)) +
  geom_path(colour = "#1d743eff", linewidth = 0.5) +
  geom_hline(yintercept = 0,             colour = "#d48282",
             linewidth = 0.4) +
  geom_hline(yintercept = meta_one$rh50, colour = "#8782d4",
             linewidth = 0.3, linetype = "dashed") +
  geom_hline(yintercept = meta_one$rh95, colour = "#8782d4",
             linewidth = 0.3, linetype = "dashed") +
  annotate("text", x = Inf, y = 0,
           label = "ground", hjust = 1.1, vjust = -0.4,
           colour = "#d48282", size = 3) +
  annotate("text", x = Inf, y = meta_one$rh50,
           label = "rh50", hjust = 1.1, vjust = -0.4,
           colour = "#8782d4", size = 3) +
  annotate("text", x = Inf, y = meta_one$rh95,
           label = "rh95 (canopy top)", hjust = 1.1, vjust = -0.4,
           colour = "#8782d4", size = 3) +
  coord_cartesian(ylim = c(-10, 120)) +
  labs(
    x = "Amplitude (noise σ units)",
    y = "Height above detected ground (m)",
    title = sprintf("Single GEDI L1B waveform — rh95 = %.0f m",
                    meta_one$rh95)
  ) +
  theme_minimal()
```

Things to notice:

- The **strong peak at 0 m** is where L2A detected ground — that's why
  `height_above_ground = 0` lines up with it. In the raw waveform this
  was at some ellipsoidal elevation; we shifted to ground-referenced
  here.
- The **broader peak around `rh95`** is the canopy top. `rh95` is
  defined so 95 % of the waveform energy (cumulative from ground up)
  has been returned by that height — you can see the waveform fall off
  sharply just above it.
- `rh50` marks the median of the return energy, usually mid-canopy.
  Between `rh50` and `rh95` is where most of the foliage signal lives.
- Anything **above `rh95`** is either sparse emergent canopy, atmospheric
  return, or noise — typically very low amplitude.

## Comparing two shots: short vs tall

The same machinery over a different shot looks very different.
Pick one shot with low `rh95` and one with high to see how canopy
structure shows up in the waveform.

```{r two-shots, eval = eval_live}
short_shot <- l1b_joined |>
  filter(rh95 < 15) |> slice(1L) |> pull(shot_number)
tall_shot  <- l1b_joined |>
  filter(rh95 > 70) |> slice(1L) |> pull(shot_number)

wf_pair <- wf |>
  filter(shot_number %in% c(short_shot, tall_shot)) |>
  left_join(
    l1b_joined |> select(shot_number, rh50, rh95),
    by = "shot_number"
  ) |>
  mutate(
    label = sprintf("shot %s\nrh50 = %.0f m, rh95 = %.0f m",
                    shot_number, rh50, rh95)
  )

ggplot(wf_pair, aes(amplitude, height_above_ground)) +
  geom_path(colour = "#1d743eff", linewidth = 0.5) +
  geom_hline(yintercept = 0, colour = "#d48282", linewidth = 0.3) +
  facet_wrap(~ label) +
  coord_cartesian(ylim = c(-10, 120)) +
  labs(
    x = "Amplitude (noise σ units)",
    y = "Height above detected ground (m)"
  ) +
  theme_minimal()
```

The short shot (open ground or low vegetation) is dominated by a single
ground peak with almost no energy above a few metres. The tall shot
has clear canopy structure: a ground return at zero plus a broad
canopy-top return at 50-90 m, possibly with understory detail in
between. Those two waveforms produced the `rh95` difference you see in
L2A — by different amounts of energy concentrating above ground.

## Composite view across the bbox

Now that you know what one waveform looks like, stack them all. We
overlay the amplitude quantiles at each height so the composite reads as
a structural profile rather than spaghetti.

```{r composite, eval = eval_live}
wf_quantiles <- wf |>
  filter(height_above_ground >= 0, height_above_ground <= 200) |>
  mutate(h_bin = round(height_above_ground)) |>
  group_by(h_bin) |>
  summarise(
    q10 = quantile(amplitude, 0.10, na.rm = TRUE),
    q50 = median(amplitude,         na.rm = TRUE),
    q90 = quantile(amplitude, 0.90, na.rm = TRUE),
    .groups = "drop"
  )

ggplot() +
  geom_path(
    data = wf,
    aes(amplitude, height_above_ground, group = shot_number),
    alpha = 0.05, colour = "#1d743eff", linewidth = 0.1
  ) +
  geom_ribbon(
    data = wf_quantiles,
    aes(xmin = q10, xmax = q90, y = h_bin),
    fill = "#1d743eff", alpha = 0.25
  ) +
  geom_path(
    data = wf_quantiles,
    aes(q50, h_bin),
    colour = "#0d4723ff", linewidth = 0.7
  ) +
  geom_hline(yintercept = 0, colour = "#d48282", linewidth = 0.3) +
  coord_cartesian(ylim = c(0, 200)) +
  labs(
    x = "Amplitude (noise σ units)",
    y = "Height above detected ground (m)",
    title    = "L1B waveforms over PNW coastal redwood",
    subtitle = "Individual shots (green paths), 10-90% amplitude range (ribbon), median (line)"
  ) +
  theme_minimal()
```

The **median line** is the "typical" waveform for this bbox: a sharp
peak at ground, a secondary peak near median canopy height, decaying
returns above it. The **10-90 % ribbon** shows shot-to-shot variation
— narrow bands mean consistent structure across the bbox, wide bands
mean heterogeneous canopy (edge shots, gaps, short vs tall mixed).
For our small redwood bbox the ribbon is tight: the forest is
structurally similar everywhere we looked.

## Cross-check: compute your own RH metrics

RH*N* is the elevation at which *N*% of cumulative waveform energy (from
ground up) has been returned. With the waveforms zeroed on ground, we can
compute it in-place and compare against L2A.

```{r rh-custom, eval = eval_live}
# cumulative energy from ground up, per shot
rh_custom <- wf |>
  filter(height_above_ground >= 0) |>       # ignore below-ground noise
  group_by(shot_number) |>
  arrange(height_above_ground, .by_group = TRUE) |>
  mutate(cum_energy = cumsum(amplitude) / sum(amplitude)) |>
  summarise(
    rh50_ours = approx(cum_energy, height_above_ground, xout = 0.50)$y,
    rh95_ours = approx(cum_energy, height_above_ground, xout = 0.95)$y,
    .groups = "drop"
  )

comparison <- l1b_joined |>
  select(shot_number, rh50, rh95) |>
  inner_join(rh_custom, by = "shot_number")

ggplot(comparison, aes(rh95, rh95_ours)) +
  geom_abline(colour = "#d48282") +
  geom_point(alpha = 0.4) +
  coord_equal() +
  labs(x = "L2A rh95 (m)", y = "Custom rh95 (m)")
```

If the points hug the 1:1 line, our simple-minded cumulative-energy RH95
agrees with L2A's more principled version. Disagreements identify shots
where L2A's noise thresholding, smoothing, or algorithm selection
(`selected_algorithm`) produces a meaningfully different answer than a
naive cumsum. Those disagreements are usually *L2A doing the right thing*,
not the other way around.

## What we skipped (and why)

- **Waveform smoothing.** L2A applies a Gaussian kernel before peak detection.
  We used raw samples. Smoothing before RH computation typically nudges
  percentile heights by <1 m but can matter in noisy shots.
- **Algorithm selection.** L2A ships six algorithm variants per shot and
  picks one via `selected_algorithm`. Our cumulative-energy RH uses none of
  that machinery.
- **Ground detection from scratch.** We used L2A's `elev_lowestmode` rather
  than detecting ground ourselves. Home-rolled ground detection in dense
  canopy or steep terrain is substantially worse than L2A's multi-year
  algorithm; when you need the ground peak, read L2A.

## Further reading

- [Hofton et al. (2019), GEDI L1B ATBD](https://lpdaac.usgs.gov/documents/982/GEDI01_B_ATBD_v2.0.pdf) — full waveform capture and calibration
- [Hofton & Blair (2019), GEDI L2A ATBD](https://lpdaac.usgs.gov/documents/588/GEDI_FCCSD_ATBD_v1.0.pdf) — peak detection and RH metric derivation
- [`sl_extract_waveforms()`] in this package — the primary helper used here
